Molecular dynamics simulation of synchronization in driven particles

### 2021-01-16 manuscript/manuscript_ajp_2020.tex
removed all \bar

%because
The inertia of 
small particles is reduced by interactions
with fluid particles \cite{Purcell1977}.
Our model 
colloids are overdamped
-- i.e. suspended in a continuous viscous fluid
that dissipates energy at a rate such 
that the particles do not accelerate.
%that greatly impedes their motions.
Energy dissipation from the fluid is modeled
with a frictional force on particle with velocity $\vec{v}_i$
%that removes energy from
%the particles,
%which we model as
$\vec{F}^{drag}_i = -\eta \vec{v}_i$
%proportional to the particle velocity 
with a friction coefficient $\eta$
proportional to the fluid viscosity.
We discuss friction models for
spheres moving through fluids in 
Section~\ref{sec:problems} Exercise~\ref{ex:reynolds}. %for more information
%We model the particle dynamics  with an overdamped equation of motion. Models with inertia set to zero are appropriate for moving through a high viscosity fluid
Newton's second law for an individual particle
%$m \vec{a}_i = \sum \vec{F}$
is simplified
by the assumption $\vec{a}_i$ is zero. % in the overdamped model.
%Newton's second law can be rearranged %for the velocity
%to
The overdamped equation of motion for an isolated particle is
%solved for velocity 
\begin{equation}
  \eta \vec{v}_i = \vec{F}^l_{i} + \vec{F}^{d}(t).
    \label{eq:motion}
\end{equation}
with  $\eta = 1$.  
%which we describe in more detail in Table~\ref{tab:1}
%and Exercise~\ref{ex:reynolds}.
In %Section~\ref{sec:problems}
Exercise~\ref{ex:n2l}
we invite the reader to confirm this result.

The equation of motion provides a direct calculation of the velocity
of an individual particle from its location $\vec{r}_i$ %in the simulation
and the simulation time.
The molecular dynamics simulation is controlled by a $for()$ loop
 which runs from an initial to maximum integer number of time steps $t_i$.
Each integer time step interval $t_{i+1}-t_i$
represents a simulation time interval of $\Delta t$ 
simulation units.  %as described in Table 1
At each time step
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then integrate
the equation of motion to move particles
to an updated position
$\vec{r}_i(t+\Delta t)$.
%
Since the acceleration as zero,
the integration of the equation of motion
is performed via 
the Euler method 
\begin{equation}
  \vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t)
    \label{eq:euler}
\end{equation}
for a small time step $\Delta t$.
In Section~\ref{sec:problems} Exercise~\ref{ex:euler}
we elaborate on
the numerical methods for 
solving differential equations,
and demonstrate how the Verlet method
simplifies to the Euler method when $\vec{a}=0$.
%
The units of the simulated variables are summarized in Table ~\ref{tab:1}.

\begin{table}[h!]
\centering
\caption{Simulation units and approximate values
  in experiment.}
\begin{ruledtabular}
\begin{tabular}{c c c } %p{5cm}}
Quantity & Simulation Unit & Experimental range\\
\hline
length &  $a_0 = 1$ & $ < a_0 <$\\
energy & $E_0 = 1$ & $ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & $<F_0<$\\
velocity &  $ a_0 / \tau $ &  $ < v < $\\
viscosity coefficient & $\eta = F_0 \tau / a_0 $ & $ <\eta<$\\
time &  $ \tau = \eta a_0 / F_0 $ & $ < \tau <$\\

@@

In Fig.~\ref{fig:0} 
the driven particle adjusts
its motion to traverse
the periodic landscape with the same rhythm as 
the applied time-dependent force $F^d(t)$.
%
In Fig.~\ref{fig:0}(a)
we plot $F^d(t)$ %defined in Eq.~\ref{eq:drive}.
as a function of time with
%The time independent
constants $F^{dc}=0.1$, 
$F^{ac}=0.05$ and $f=0.01$ cycles per time unit.
The temporal period of the driving force is
$T = 1/f = 100$ time units.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%RAMP
%To avoid transient oscillations at early times, the dc driving force is initially zero and slowly increased %$F^{dc}$. at a rate of $\Delta F^{dc} = 0.01$ every $\Delta t = 4$ time units  to the constant value $F^{dc}=0.1$. In Fig.~\ref{fig:0}(a) we plot $F^{dc}$ in blue to show the increase in  $F^{dc}$ for the first $40$ time units In orange we show  the combined ac and dc drive $F^d(t)$ when the magnitude of the ac applied force is initially zero.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%since $\sin{(0)}=0$.
We mark the maxima of $F^{ac}(t)$
with the
vertical dashed lines %in panels (a) and (b)
-- i.e. $\sin{(\omega t)} = 1$.

@@

In Fig.~\ref{fig:0}(b) 
we show the $y-$position of the particle
as a function of time.
We 
normalize $y$ by $\lambda$, the spatial period of the potential
to show how the particle moves between substrate minima.  
The initial particle position is $y = 1$,
where the $F^l$ is negative, 
causing the particle to move to
a landscape minima at the start of the simulation.
As the driving force increases,
%the effects of the landscape become less obvious,
the motion of the particle
becomes locked to the driving period $T$.
The particle moves
in the positive y-direction
through $\Delta y = 2 \lambda$ in one time period $\Delta t = T$,
with 
the average velocity 
$\bar{v}_y = 2 \lambda f$. % where $T = 1/f$.
The horizontal dashed lines
coincide with the condition the substrate force is
minimum $F^l = -A_p$,  
i.e. $\sin{(\pi y / \lambda)} = 1$.
%or $\pi y / \lambda = m \pi/2$, where $m = 1, 5, 7$, {\it etc.}
Once the particle motion
coincides with the applied drive, 
the particle position $y$
crosses the intersection of
the horizontal and vertical lines.
This indicates 
the driving force is maximum positive when the
force on the particle from the landscape is maximum negative,
indicating synchronization.
The slope of $y/\lambda$
is proportional to the net force on the particle.
This can be seen in 
Fig.~\ref{fig:0} %we demonstrate this relationship,
where local extrema appear in $y/\lambda$
when $F_{D}(t) \approx 0 $. 
We invite interested students to repeat this
simulation for any initial particle position %of $y$
and a range of simulation constants
to demonstrate that synchronization in this model
does not depend on a particular set of conditions.
%but is a direct consequence of the simplicity of the model.

@@

The competition between the driving force and landscape potential
can produce a variety of hopping patterns in the particle motion. 
%In Fig.~\ref{fig:0}
%reaching a minimum position
%of $y/\lambda = 1$
%over the total period $T = 100$.
%In Fig. ~\ref{fig:0}
%the average displacement during one time period is a single wavelength
%of the substrate $y(t_0+T) - y(t_0) = \lambda$.
%single dc case with figure
%dc_current 0.01
%ac_current 0.05
%ac_frequency 0.01
%fp_y 0.1
%
The relative values of $F^{ac}$, $F^{dc}$ and $A_p$
control the rate and distance a  particle moves 
forward and backward in the landscape.
To explore the possible hopping patterns,
we sweep through a range of $F^{dc}$ for fixed $F^{ac}$ and $A_p$,
as shown in Fig.~\ref{fig:1}(b).
We keep $F^{dc}$ constant
for $\Delta t = 10^5$ simulation timesteps
and 
measure the average velocity $\langle v_y \rangle $ 
as a function of $F^{dc}$.
We plot the results in Fig.~\ref{fig:1}
for $F^{ac} = 0.05$, $f=0.01$, and the landscape force $\vec{F}^{l}$
shown in Fig.~\ref{fig:0}.
The steps in $\langle v_y \rangle$ occur because 
the average velocity of the particle
is constant across a range of $F^{dc}$.
This synchronization
of the particle velocity can be measured
in terms of the landscape period.
%despite the overall magnitude
%of $F^{dc}$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle.
The average velocity $\langle{v}_y\rangle$
is the average displacement $\Delta y = y(t_0+T) - y(t_0)$
over the period of the driving force.

If $A_p$ is large compared to the extrema of $F^{d}(t)$,
the particle oscillates back and forth
in a single minima
with no net velocity.
In Fig.~\ref{fig:1}, this occurs at low
low $F^{dc}$ when $\langle v_y \rangle$ is zero.

The particle 
achieves a variety of oscillation modes as a function of $F^{dc}$.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F^{dc}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}.
Here $\langle v_{y} \rangle$ is increasing in non-uniform steps,
with a quantized height of
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.  
%$\lambda = L/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
%of the landscape,
%and $f = 0.01$ cycles per time unit.

@@

\item Newton's second law states that
  the acceleration of a particle $i$
  is proportional to 
  the sum of forces on a particle 
  \begin{equation}
  m_i \vec{a}_i = \sum \vec{F}_i
  \label{eq:n2l}
  \end{equation}
  where the constant of proportionality is the
  inertial mass $m_i$
  The addition of a dissipative force in a dynamical model
  on a colloid is typically modeled
  with a drag force proportional to the particle's velocity
  in the opposite direction of motion 
  $\vec{F}^{drag} = - \eta \vec{v}_i$
  where $\eta$ is the drag coefficient.
  Demonstrate
  the ratio of $m/\eta$ has units of time, 
  %This ratio of the acceleration coefficient $m$
  %to the drag coefficient $\eta$
  %is
  known as the momentum relaxation time.
  The momentum relaxation time is 
  known to be small for
  particles with low Reynolds numbers,
  indivating 
  a regime in which the acceleration term can be ignored
  entirely.
  Demonstrate that a particle confined to a landscape exerting force
  $F^l(\vec{r}_i)$ subject to a time dependent drive $F^d(t)$
  can be modeled with the equation of motion described in 
  Eq.~\ref{eq:motion}. 
  \label{ex:n2l}

\item
  Generally when solving differential equations,
  the Euler method is effective for solving linear equations,
  such as the one-dimensional 
  $f(y) = df/dy$ 
  but not higher order methods.

@@

@@

@@

\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{twenty}
\caption{\textbf{(a)} A single particle (colored orange - mark in some manner for non-color views) is driven with a constant amplitude $F^{ac}$ and frequency $\omega$ through 19 neighboring particles (colored green - mark differently) confined by a quasi one-dimensional channel. The landscape is colored as in Fig.~\ref{fig1}(a). \textbf{(b)} Average $\bar{v}_{y}$ versus $F^{dc}$, where $\bar{v}_{y}$ is the average particle velocity of the driven particle in the y-direction.}
\label{fig:2}
\end{figure}
\end{center}

>>>>

%because
The inertia of 
small particles is reduced by interactions
with fluid particles \cite{Purcell1977}.
In this model 
colloids are overdamped
-- i.e. suspended in a continuous viscous fluid
that dissipates energy %at a rate
such 
that the particles do not accelerate.
%that greatly impedes their motions.
Energy dissipation from the fluid is modeled
with a frictional force on particle with velocity $\vec{v}_i$
%that removes energy from
%the particles,
%which we model as
$\vec{F}^{drag}_i = -\eta \vec{v}_i$
%proportional to the particle velocity 
with a friction coefficient $\eta$
proportional to the fluid viscosity.
We discuss friction models for
spheres moving through fluids in 
%Section~\ref{sec:problems}
Exercise~\ref{ex:reynolds}. %for more information
%We model the particle dynamics  with an overdamped equation of motion. Models with inertia set to zero are appropriate for moving through a high viscosity fluid
Newton's second law for an individual particle
%$m \vec{a}_i = \sum \vec{F}$
is simplified
by the assumption $\vec{a}_i$ is zero. % in the overdamped model.
%Newton's second law can be rearranged %for the velocity
%to
The overdamped equation of motion for an isolated particle is
%solved for velocity 
\begin{equation}
  \eta \vec{v}_i = \vec{F}^l_{i}(\vec{r}_i) + \vec{F}^{d}(t).
    \label{eq:motion}
\end{equation}
with  $\eta = 1$.  
%which we describe in more detail in Table~\ref{tab:1}
%and Exercise~\ref{ex:reynolds}.
%In %Section~\ref{sec:problems}
%Exercise~\ref{ex:n2l}
%we invite the reader to confirm this result.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location $\vec{r}_i$ %in the simulation
and the simulation time.

The molecular dynamics simulation is controlled by a $for()$ loop
which runs from an initial to maximum integer number of time steps.
%$t_i$.
Each integer time step interval %$t_{i+1}-t_i$
represents a simulation time interval of $\Delta t$ 
simulation units.  %as described in Table 1
At each time step
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then integrate
the equation of motion to move particles
to an updated position.
%$\vec{r}_i(t+\Delta t)$.%
Since the acceleration as zero,
the integration of the equation of motion
is performed via 
the Euler method 
\begin{equation}
  \vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t)
    \label{eq:euler}
\end{equation}
for a small time step $\Delta t = 0.001 \tau$.
In %Section~\ref{sec:problems}
Exercise~\ref{ex:euler}
we describe 
the numerical methods for 
solving differential equations.
%
The units of the simulated variables are summarized in Table ~\ref{tab:1}.

\begin{table}[h!]
\centering
\caption{Simulation units and approximate values
  in experiment. (coming soon!)}
\begin{ruledtabular}
\begin{tabular}{c c c } %p{5cm}}
Quantity & Simulation Unit & Experimental range\\
\hline
length &  $a_0 = 1$ & $ < a_0 <$\\
energy & $E_0 = 1$ & $ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & $<F_0<$\\
velocity &  $ a_0 / \tau $ &  $ < v < $\\
viscosity coefficient & $\eta = F_0 \tau / a_0 $ & $ <\eta<$\\
time &  $ \tau = \eta a_0 / F_0 $ & $ < \tau <$\\

@@

In Fig.~\ref{fig:0} 
the driven particle adjusts
its motion to traverse
the periodic landscape with the same rhythm as 
the applied time-dependent force $F^d(t)$.
%
In Fig.~\ref{fig:0}(a)
we plot $F^d(t)$ %defined in Eq.~\ref{eq:drive}.
as a function of time with
%The time independent
constants $F^{dc}=0.1$, 
$F^{ac}=0.05$ and $f=0.01$ cycles per time unit $\tau$.
The temporal period of the driving force is
$T = 1/f = 100 \tau$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%RAMP
%To avoid transient oscillations at early times, the dc driving force is initially zero and slowly increased %$F^{dc}$. at a rate of $\Delta F^{dc} = 0.01$ every $\Delta t = 4$ time units  to the constant value $F^{dc}=0.1$. In Fig.~\ref{fig:0}(a) we plot $F^{dc}$ in blue to show the increase in  $F^{dc}$ for the first $40$ time units In orange we show  the combined ac and dc drive $F^d(t)$ when the magnitude of the ac applied force is initially zero.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%since $\sin{(0)}=0$.
We mark the maxima of $F^{ac}(t)$
with the
vertical dashed lines %in panels (a) and (b)
-- i.e. $\sin{(\omega t)} = 1$.

@@

In Fig.~\ref{fig:0}(b) 
we show the $y-$position of the particle
as a function of time.
We 
normalize $y$ by $\lambda$, the spatial period of the potential
to show how the particle moves between substrate minima.  
The initial particle position is $y = 1 a_0$,
where the $F^l$ is negative.
%JUST BECAUSE
%causing the particle to move to
%a landscape minima at the start of the simulation.
As the driving force increases,
%the effects of the landscape become less obvious,
the motion of the particle
synchronizes with the driving period $T$.
The particle moves
in the positive y-direction
through $\Delta y = 2 \lambda$ over the time period $T$,
with 
the average velocity 
$\langle {v}_y \rangle= 2 \lambda f$. % where $T = 1/f$.
The horizontal dashed lines
coincide with the condition the substrate force is
minimum $F^l = -A_p$,  
i.e. $\sin{(\pi y / \lambda)} = 1$.
%or $\pi y / \lambda = m \pi/2$, where $m = 1, 5, 7$, {\it etc.}
%Once the particle motion
%coincides with the applied drive,
%indicating synchronization,
The %shows %indicates the
driving force is maximum when the landscape 
force on the particle is minimum,
as indicated with the coincidence of 
the particle position $y$
with the intersection of
the horizontal and vertical lines.
The slope of $y/\lambda$
is proportional to the net force on the particle.
This can be seen in 
Fig.~\ref{fig:0} %we demonstrate this relationship,
where local extrema appear in $y/\lambda$
when $F_{D}(t) \approx 0 $.

@@

The competition between the driving force and landscape potential
can produce a variety of hopping patterns in the particle motion. 
%In Fig.~\ref{fig:0}
%reaching a minimum position
%of $y/\lambda = 1$
%over the total period $T = 100$.
%In Fig. ~\ref{fig:0}
%the average displacement during one time period is a single wavelength
%of the substrate $y(t_0+T) - y(t_0) = \lambda$.
%single dc case with figure
%dc_current 0.01
%ac_current 0.05
%ac_frequency 0.01
%fp_y 0.1
%
The relative values of $F^{ac}$, $F^{dc}$ and $A_p$
control the rate and distance a  particle moves 
forward and backward in the landscape.
To explore the possible hopping patterns,
we sweep through a range of $F^{dc}$ for fixed $F^{ac}$ and $A_p$,
as shown in Fig.~\ref{fig:1}(b).
We keep $F^{dc}$ constant
for $10^5$ simulation timesteps
and 
measure the average velocity $\langle v_y \rangle $ 
as a function of $F^{dc}$.
The average velocity $\langle{v}_y\rangle$
is the average displacement %$\Delta y = y(t_0+T) - y(t_0)$
over the period of the driving force.
We plot the results in Fig.~\ref{fig:1}
for $F^{ac} = 0.05$, $f=0.01$, and the landscape %force $\vec{F}^{l}$
shown in Fig.~\ref{fig:0}.
Discrete steps in $\langle v_y \rangle$ occur because 
the average velocity of the particle
is constant across a range of $F^{dc}$.
This synchronization
of the particle velocity can be measured
in terms of the landscape period.
%despite the overall magnitude
%of $F^{dc}$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle.

If $A_p$ is large compared to the extrema of $F^{d}(t)$,
the particle oscillates back and forth
in a single minima
with no net velocity.
In Fig.~\ref{fig:1}, this occurs at 
low $F^{dc}$ when $\langle v_y \rangle$ is zero.

The particle 
achieves a variety of oscillation modes as a function of $F^{dc}$.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\langle {v}_{y} \rangle$
over a range of driving forces $F^{dc}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}.
Here $\langle v_{y} \rangle$ is increasing in non-uniform steps,
with a quantized height of
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.  
%$\lambda = L/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
%of the landscape,
%and $f = 0.01$ cycles per time unit.

@@

\item{Equation of Motion and Integration Methods}

  Newton's second law states that
  the acceleration of a particle $i$
  is proportional to 
  the sum of forces on a particle 
  \begin{equation}
  m_i \vec{a}_i = \sum \vec{F}_i
  \label{eq:n2l}
  \end{equation}
  where the constant of proportionality is the
  inertial mass $m_i$
  The addition of a dissipative force in a dynamical model
  on a colloid is typically modeled
  with a drag force proportional to the particle's velocity
  in the opposite direction of motion 
  $\vec{F}^{drag} = - \eta \vec{v}_i$
  where $\eta$ is the drag coefficient.
  Demonstrate
  the ratio of $m/\eta$ has units of time, 
  %This ratio of the acceleration coefficient $m$
  %to the drag coefficient $\eta$
  %is
  known as the momentum relaxation time.
  The momentum relaxation time is 
  known to be small for
  particles with low Reynolds numbers,
  indivating 
  a regime in which the acceleration term can be ignored
  entirely.
  Demonstrate that a particle confined to a landscape exerting force
  $F^l(\vec{r}_i)$ subject to a time dependent drive $F^d(t)$
  can be modeled with the equation of motion described in 
  Eq.~\ref{eq:motion}. 
  %\label{ex:n2l}

%\item
  Generally when solving differential equations,
  the Euler method is effective for solving linear equations,
  such as the one-dimensional 
  $f(y) = df/dy$ 
  but not higher order methods.

@@

  and demonstrate how the Verlet method
  simplifies to the Euler method when $\vec{a}=0$.

@@

  We invite interested students to repeat this
simulation for any initial particle position %of $y$
and a range of simulation constants
to demonstrate that synchronization in this model
does not depend on a particular set of conditions.
%but is a direct consequence of the simplicity of the model.

@@

\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{twenty}
\caption{\textbf{(a)} A single particle (colored orange - mark in some manner for non-color views) is driven with a constant amplitude $F^{ac}$ and frequency $\omega$ through 19 neighboring particles (colored green - mark differently) confined by a quasi one-dimensional channel. The landscape is colored as in Fig.~\ref{fig1}(a). \textbf{(b)} Average $\langle {v}_{y} \rangle $ versus $F^{dc}$, where $\langle {v}_{y} \rangle$ is the average particle velocity of the driven particle in the y-direction.}
\label{fig:2}
\end{figure}
\end{center}
